Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
cutting

\section{Experimental Validation}
\label{sec:expe}
In this section, we describe a series of numerical experiments to
confirm the performance predictions of Section~\ref{sec:model}. To
better understand our results, we first perform some bandwidth
measurements, followed by the actual SpMV computations.

In this experimental section, as a point of reference we also provide
some results on a typical CPU architecture. This system is
...\todo{include a short description of the host.}  All the codes are
compiled with icc xxx and -O3 optimization. The MIC codes use the IMCI
instructions while the CPU codes use AVX instructions.

@@

\begin{figure*}[bth]
  \centering
  \subfigure[On MIC with 240 threads, 64-byte memory
    alignment.]{\includegraphics[width=.40\linewidth]{figures/bandwidth_read_write.pdf}\label{fig:band_rw_MIC}}
  %
  \subfigure[On CPU with 32 threads, 32-byte memory
    alignment.]{\includegraphics[width=.40\linewidth]{figures/host_test1_readwrite_no_temporal_hint.pdf}\label{fig:band_rw_CPU}}

  \caption{Bandwidth performance on the host under idealized
    conditions as a function of matrix row size.  Entries with "cpp"
    denote cases where coding was performed without MIC vector
    instructions.  \todo{Can we label the x-axis "number of floats" or
      "size in MB"?}\todo{Can we use the same symbols in both charts?}
    \todo{Can we change the color scheme? in (a), I do not see the
      difference between "read\_cpp" and "read/write cpp"} }
  \label{fig:bandwidth}
\end{figure*}

@@

Looking at the results on the host in Figure~\ref{fig:band_rw_CPU}
show no significant difference between manual vectorization and the
compiler generated one (there is no store nrngo on the host). It also
shows quite stable performance with about 70GB/s in the {\tt read}
benchmark, about 40GB/s in the {\tt write} benchmark and about 49 in
the {\tt read write} benchmark. The bandwidth achieved on MIC is about
2 to 3 times faster than the bandwidth achieved on the CPU. The
kernels we consider being memory bound, we expect MIC to be 2 to 3
times faster.

@@

To summarize, using vector instruction seems mandatory to reach the
highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140GB/s to 160GB/s. That is why we set in our estimation the best
achievable bandwidth to 150GB/s despite carefully crafted code to
maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}. Comparing to a CPU architecture, we
expect MIC to be 2 to 3 times faster.

@@

\begin{figure*}
  \centering \subfigure[Performance of $y=Ax$ on the MIC. Squares:
    base 1/1 case, solid circles: 4/4 case implemented in C++, solid
    triangles: 4/4 case implemented with MIC vector instructions. Each
    symbol has two cases: with (red) and without (blue) Reverse
    Cuthill-McKee bandwidth reduction. The grid is $96^3$]
             {\includegraphics[width=.40\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
\hspace{.4in}
%
  \subfigure[Performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.40\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }\todo{legend}}
  \label{fig:comparison_spmv}
\end{figure*}

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of 4 vectors by 4 matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
4 vectors by 4 matrices using either standard C++ code and our
optimized implementation. On MIC (Figure~\ref{fig:perf_mic}) the
classical SpMV computation reaches 15Gflop/s\todo{check}. The standard
C++ implementation reaches a performance of 40Gflop/s while our
optimized implementation reaches 137Gflop/s\todo{check}. Notice that
RCM ordering has no impact in the standard C++ implementation while it
provides an significant improvement in our manually vectorized
version. This highlights the C++ implementation is instruction bound
while our manually vectorized implementation is most likely memory
bound.  As a point of comparison, we show an AVX implementation on the
CPU of the two kernels and we show the best achievable performance in
the classical SpMV in Figure~\ref{fig:perf_host}. \todo{waiting for
  complete plot before writing more.}

>>>>

\section{Experimental Validation}
\label{sec:expe}
In this section, we describe a series of numerical experiments to
confirm the performance predictions of Section~\ref{sec:model}. To
better understand our results, we first perform some bandwidth
measurements, followed by the actual SpMV computations. All the codes
are compiled with the Intel C Compiler version xxx\todo{gordon?} and
-O3 optimization. The MIC codes use the IMCI instructions exposed by
the compiler through intrinsics.

@@

\begin{figure}[bth]
  \centering
  \includegraphics[width=.9\linewidth]{figures/bandwidth_read_write.pdf}

  \caption{On MIC with 240 threads, 64-byte memory
    alignment. Bandwidth performance under idealized
    conditions as a function of matrix row size.  Entries with "cpp"
    denote cases where coding was performed without MIC vector
    instructions. \todo{legend}  \todo{Can we label the x-axis "number of floats" or
      "size in MB"?}\todo{Can we use the same symbols in both charts?}
    \todo{Can we change the color scheme? in (a), I do not see the
      difference between "read\_cpp" and "read/write cpp"} }
  \label{fig:bandwidth}
  \label{fig:band_rw_MIC}
\end{figure}

@@

@@

To summarize, using vector instruction seems mandatory to reach the
highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140GB/s to 160GB/s. That is why we set in our estimation the best
achievable bandwidth to 150GB/s despite carefully crafted code to
maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}.

@@

\begin{figure}
  \centering

  \includegraphics[width=.9\linewidth]{figures/mic_performance_nb_threads.pdf}

  \caption{Performance of $y=Ax$ on the MIC. Squares:
    base 1/1 case, solid circles: 4/4 case implemented in C++, solid
    triangles: 4/4 case implemented with MIC vector instructions. Each
    symbol has two cases: with (red) and without (blue) Reverse
    Cuthill-McKee bandwidth reduction. The grid is $96^3$. Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{legend} \todo{fix caption}}
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of 4 vectors by 4 matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
4 vectors by 4 matrices using either standard C++ code and our
optimized implementation. The
classical SpMV computation reaches 15Gflop/s\todo{check}. The standard
C++ implementation reaches a performance of 40Gflop/s while our
optimized implementation reaches 137Gflop/s\todo{check}. Notice that
RCM ordering has no impact in the standard C++ implementation while it
provides an significant improvement in our manually vectorized
version. This highlights the C++ implementation is instruction bound
while our manually vectorized implementation is most likely memory
bound.
